Pre working group data exploration: Mainly wrt coral cover

### 2024-07-04 julianna_data_exploration.Rmd
Did a bunch of exploration to figure out what the bottom bound structure is in the Shannon algal vs. coral/turf/algae plots. Believe it has to do with instances when you have one dominant thing (e.g., turf algae) and then increase the proportion of a second alga incrementally at 4% intervals

```{r}
allbenthos <- read_csv(here("data", "allbenthicdata_cleaned_colnames.csv")) 
algae <- read_csv(here("data", "algaeonly_cleaned_colnames.csv"))
```

@@

```{r}
algae %>% 
  # grab the metadata info that is in the same order (can re-join after)
  select(year, location, site, habitat, site_habitat, transect, quadrat) %>% 
  mutate(Richness = specnumber(sppMat),
         Shannon = diversity(sppMat, "shannon"),
         Richness = specnumber(sppMat)) -> algDiv
```

@@

@@

```{r}
allbenthos %>% 
  select(year, location, site, habitat, site_habitat, transect, quadrat, coral) %>% 
  full_join(algDiv) %>% 
  # add column for each transect ID
  rowwise() %>% 
  mutate(transect_id = paste0(site, habitat, year, transect)) -> coralAlgDiv  
```

## Plot shannon

@@

```{r}
coralAlgDiv %>% 
  ggplot(aes(x = coral, y = Shannon, color = year)) +
  geom_point() + 
  facet_grid(habitat~site) +
  theme_classic()
```

@@

@@

```{r}
m1 <- glmer(Richness ~ scale(coral) + habitat + (1|site) + (1|transect_id) + (1|year), 
            family = poisson(link = "log"), data = coralAlgDiv)
  plot(m1)
  summary(m1)
  emmeans(m1, ~ habitat)
```

>>>>

```{r}
allbenthos <- read_csv(here("data", "allbenthicdata_cleaned_colnames.csv")) 
algae <- read_csv(here("data", "algaeonly_cleaned_colnames.csv")) %>% 
  # going to run everything without algal turf
  select(-algal_turf)
```

@@

```{r}
algae %>% 
  # grab the metadata info that is in the same order (can re-join after)
  select(year, location, site, habitat, site_habitat, transect, quadrat) %>% 
  mutate(Richness = specnumber(sppMat),
         Shannon = diversity(sppMat, "shannon"),
        # Evenness = diversity(sppMat, "ev")
         Richness = specnumber(sppMat)) -> algDiv
```

@@

# Compare algal and coral cover

```{r}
# get ranges to calculate over
startAlg <- grep("acanthophora_spicifera", colnames(algae))
endAlg <- grep("valonia_ventricosa", colnames(algae))

algae %>% 
  mutate(AlgSum = rowSums(algae[startAlg:endAlg])) -> algSum

algSum %>% 
  full_join(allbenthos) %>% 
  ggplot(aes(x = coral, y = AlgSum, color = year)) +
  geom_point() +
  facet_grid(site~habitat)

@@

```{r}
allbenthos %>% 
  select(year, location, site, habitat, site_habitat, transect, quadrat, coral) %>% 
  full_join(algDiv) %>% 
  # add column for each transect ID
  rowwise() %>% 
  mutate(transect_id = paste0(site, habitat, "_", year, "_", transect)) -> coralAlgDiv  
```

## Plot Shannon

@@

```{r}
coralAlgDiv %>% 
  ggplot(aes(x = coral, y = Shannon, color = as.factor(Richness))) +
  geom_point(alpha = 0.5) + 
  facet_grid(habitat~site) +
  theme_classic() + 
  geom_hline(yintercept = 0.693, color = "red", linetype = "dashed") # this is when two species are equally distributed (half half)--not sure why that would be
```

@@

Or look at Shannon vs. the percent cover of algae in the plot

```{r}
coralAlgDiv %>% 
  full_join(algSum) %>% 
  ggplot(aes(x = AlgSum, y = Shannon, color = as.factor(Richness))) +
  geom_point(alpha = 0.5) + 
  facet_grid(habitat~site) +
  theme_classic() + 
  geom_hline(yintercept = 0.693, color = "red", linetype = "dashed")
```

What about this when Richness == 2?

```{r}
coralAlgDiv %>% 
  full_join(algSum) %>% view()
  filter(Richness == 2) %>% 
  ggplot(aes(x = AlgSum, y = Shannon, color = as.factor(Richness))) +
  geom_point(alpha = 0.5) + 
  facet_grid(habitat~site) +
  theme_classic() + 
  geom_hline(yintercept = 0.693, color = "red", linetype = "dashed")
```

Could this be because the only way to get a high enough imbalance/unevenness is having a large enough algal N

What does it look like relative to turf algae?

```{r}
coralAlgDiv %>% 
  full_join(algSum) %>% 
  filter(Richness == 2) %>% 
  ggplot(aes(x = algal_turf, y = Shannon, color = as.factor(Richness))) +
  geom_point(alpha = 0.5) + 
  facet_grid(habitat~site) +
  theme_classic() + 
  geom_hline(yintercept = 0.693, color = "red", linetype = "dashed")
```

Nic: are these lines all turf but 1 line, all turf but 2 points line, all turf but 3 points line?

```{r}
coralAlgDiv %>% 
  full_join(algSum) %>% 
  filter(Richness == 2) %>% 
  mutate(Nonturf = AlgSum - algal_turf) %>% select(Nonturf) %>% unique() %>% view()
  ggplot(aes(x = algal_turf, y = Shannon, color = as.factor(Nonturf))) +
  geom_point(alpha = 0.5) + 
  facet_grid(habitat~site) +
  theme_classic() + 
  geom_hline(yintercept = 0.693, color = "red", linetype = "dashed")
```

^This seems to be true here. It seems like lines have to do with how much of the non-turf you have. They only do 25 points and multiply by 4 so everything is a multiple of 4. But then how can we have a 2 there???

## Problem solve Shannon

Trying to instead calculate Shannon manually with 100 as the denominator?

```{r}
shannon100 <- function(x, equalize.groups = FALSE, base = exp(1)) {
  x <- drop(as.matrix(x))
  if (length(dim(x)) > 1) {
        x <- sweep(x, MARGIN = 1, 100, FUN = "/") # MARGIN = 1 I think means to do this rowwise; FUN = "/" means to divide by 100 (replaced the row sum with 100 here)
    }
    else {
        x <- x/(100 <- sum(x)) # replaced total with 100 here
    }
  x <- -x * log(x, base)
  if (length(dim(x)) > 1) 
        H <- apply(x, MARGIN = 1, sum, na.rm = TRUE)
    else H <- sum(x, na.rm = TRUE)
  H
}

# add this to the df
algae %>% 
  # grab the metadata info that is in the same order (can re-join after)
  select(year, location, site, habitat, site_habitat, transect, quadrat) %>% 
  mutate(Richness = specnumber(sppMat),
         Shannon = diversity(sppMat, "shannon"),
         Shannon100 = shannon100(sppMat),
         Richness = specnumber(sppMat)) -> algDiv100
```

Plot

```{r}
algDiv100 %>% 
  ggplot(aes(x = Shannon100, y = Shannon)) +
  geom_point() +
  geom_abline(intercept = 0, slope = 1, color = "red", linetype = "dashed") 
```

Re-make old plot

```{r}
# re-join with coral
allbenthos %>% 
  select(year, location, site, habitat, site_habitat, transect, quadrat, coral) %>% 
  full_join(algDiv100) %>% 
  # add column for each transect ID
  rowwise() %>% 
  mutate(transect_id = paste0(site, habitat, "_", year, "_", transect)) -> coralAlgDiv100

# plot  
coralAlgDiv100 %>% 
  ggplot(aes(x = coral, y = Shannon, color = year)) +
  geom_point() + 
  facet_grid(habitat~site) +
  theme_classic() # NOPE very much still there 
```

@@

```{r}
m1 <- glmer(Richness ~ scale(coral) + habitat + (1|site) + (1|transect_id) + (1|year), 
            family = poisson(link = "log"), data = coralAlgDiv)
  plot(m1) # I think this looks pretty weird right?
  summary(m1)
  emmeans(m1, ~ habitat)
```
